Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-12 ind_files/01_Introduction.tex
Minor phrasing changes, add some references

Chapter \ref{chap_infD_processes} introduces a function-valued process\footnote{This is really a measure-valued process but we will pretend we don't know this in the interest of accessibility} to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's infinite alleles model and Lande's gradient dynamics can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research~\citep{walsh_introduction_1986,carmona_stochastic_1999,da_prato_stochastic_2014,balan_gentle_2018} and the mathematics itself is thus still far from settled. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues~\citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, has never been presented in full generality either mathematically or biologically. Mathematically, chapter \ref{chap_infD_processes} presents heuristic, accessible alternatives to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits~\citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021} by generalizing the idea of a system size expansion of density-dependent (finite-dimensional) birth-death processes to the infinite-dimensional case using the notion of functional differentiation. Biologically, Chapter \ref{chap_infD_processes} provides a series of functional partial differential equations (`stochastic field equations') to describe the dynamics of one-dimensional quantitative traits in populations. Some concrete models are presented as examples in Appendix \ref{App_examples} for clarity regarding the major ideas.

>>>>

Chapter \ref{chap_infD_processes} introduces a function-valued process\footnote{This is really a measure-valued process but we will pretend we don't know this in the interest of accessibility} to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's infinite alleles model and Lande's gradient dynamics can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research~\citep{carmona_stochastic_1999,da_prato_stochastic_2014,prevot_concise_2007,liu_stochastic_2015,bogachev_fokker-planck-kolmogorov_2015,balan_gentle_2018} and the mathematics itself is far from settled. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues~\citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, has never been presented in full generality before. Mathematically, chapter \ref{chap_infD_processes} presents heuristic, accessible alternatives to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits~\citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021} by generalizing the idea of a system size expansion of density-dependent (finite-dimensional) birth-death processes to the infinite-dimensional case using the notion of functional differentiation. Biologically, Chapter \ref{chap_infD_processes} provides `stochastic field equations' that describe the dynamics of one-dimensional quantitative traits in finite populations and illustrates that these equations are consistent with well-known formalisms in quantitative genetics at the infinite population limit. Some concrete models are presented as examples in Appendix \ref{App_examples} for clarity regarding the major ideas.

### 2023-03-12 ind_files/04_unification_discussion.tex
Minor phrasing changes, add some references

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful, and indeed has been extensively exploited~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because many problems are much more easily attacked using probabilistic tools that operate on It\^o process, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the `physics' side through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example for studying phenotypic clustering and evolutionary branching), slow manifold (`adiabatic') approximations~\citep{parsons_dimension_2017}, and WKB style expansions~\cite{rogers_demographic_2012,assaf_wkb_2017}; though WKB theory is, of course, closely related to large deviation principles that occur in the study of Markov processes and come in a distinctly `probabilistic' flavor, see for example~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the `mathematics'\footnote{The `mathematics' vs `physics' distinction is just a caricature and is not intended to be read into or thought about too deeply :)} side by working with branching processes/SPDEs and using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators (semigroup theory), weak/mild solutions, and the martingale problem (See for example~\cite{etheridge_mathematical_2011} or~\cite{week_white_2021}). Rigorously establishing a relation between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and SPDEs of the form \eqref{functional_langevin} (or those studied in~\cite{week_white_2021}) for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object.

To the best of my knowledge, a general formulation of stochastic field equations for the population dynamics of quantitative traits from the functional Fokker-Planck side in the manner I have carried out here has not been done before. Similar equations have been formulated for some specific stochastic models of quantitative trait evolution~\citep{rogers_demographic_2012,rogers_modes_2015} and population ecology of size-structured communities~\citep{odwyer_integrative_2009}. Stochastic field equations are also known in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014}, and have recently been proposed in a model of collective motion~\citep{o_laighleis_minimal_2018}. Broadly similar deterministic field theoretic approaches have also been proposed to study bio-geography and spatial biodiversity patterns such as the species-area relationship~(\cite{odwyer_field_2010}; But see~\cite{grilli_absence_2012} for an explanation of why the calculations presented in~\cite{odwyer_field_2010} are not exactly correct\footnote{The essential problem is that an assumption about time reversibility (`detailed balance' in more accurate physics lingo), usually made for equilibrium systems, is not satisfied. Nevertheless, the general \emph{approach} is still interesting and seems reasonable}). The ubiquity of such approaches in diverse areas of theoretical biology suggests that the formal systematic study and analysis of such equations could have wide-spread applications. Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory or quantum field theory, and are largely attacked using ingenious (and often somewhat system-specific) arguments and tools that may not necessarily be mathematically well understood~\citep{carmona_stochastic_1999}. Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - For example, a careful reading of the literature provides many hints and recent attempts illustrating how the major ideas behind the Fock space representation~\citep{del_razo_probabilistic_2022}, perturbative `loop diagrams/expansions'~\citep{dodd_many-body_2009}, and the path integral formalism~\citep{doi_second_1976, peliti_path_1985, chow_path_2015, weber_master_2017} can all be co-opted to coax biological insights from the sort of stochastic processes I formulate as a model for evolutionary dynamics via equation \eqref{disc_functional_field_eqns}.

Equation \eqref{disc_functional_field_eqns} thus provides a more accessible (to some) alternative formulation that does not (explicitly) rely on tools from relatively `advanced' mathematical fields such as martingale theory and stochastic analysis that are typically used to study the evolution of quantitative traits in finite, fluctuating populations~\citep{dawson_stochastic_1975,fleming_measure-valued_1979,ethier_markov_1986,champagnat_unifying_2006,etheridge_mathematical_2011,week_white_2021}. This is intended to make the models and ideas more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis and encourage the use of techniques from physics and related domains, as outlined in the paragraphs above. Importantly, the formalism I develop here in terms of functional Fokker-Planck equations likely does \emph{not} carry over to the study of higher dimensional quantitative traits (or populations which vary in two or more one-dimensional quantitative traits) because these processes are routinely badly behaved in higher dimensions: In particular, a probability density $P(\phi(x),t)$ with respect to the reals frequently does not even exist in higher dimensions if one has any biologically non-trivial features such as interactions between types (\cite{fleming_measure-valued_1979, walsh_introduction_1986}; Also see for example~\cite{evans_measure-valued_1994}), rendering equation \eqref{disc_functional_field_eqns} entirely meaningless. My (admittedly limited) understanding is that such processes are also rather difficult to study in two or more dimensions from the SPDE side for similar reasons since one is forced to work with distribution valued (rather than function valued) solutions and thus needs to use a considerable amount of advanced functional analysis to establish even basic existence-uniqueness results~\citep{walsh_introduction_1986,carmona_stochastic_1999,balan_gentle_2018}. It may well be the case that concrete biologically useful progress in this direction requires new mathematics altogether, a situation increasingly also encountered in other areas of mathematical biology~\citep{vittadello_open_2022}. Much in the way physics has historically led to development in mathematics~\citep{cohen_mathematics_2004}, the pursuit of general biological theory may therefore inspire new ideas that enrich and enliven `pure' mathematics.

>>>>

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful, and indeed has been extensively exploited~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because many problems are much more easily attacked using probabilistic tools that operate on It\^o process, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the `physics' side through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example for studying phenotypic clustering and evolutionary branching), slow manifold (`adiabatic') approximations~\citep{parsons_dimension_2017}, and WKB style expansions~(\cite{rogers_demographic_2012,assaf_wkb_2017}; though it bears noting that WKB theory is closely related to large deviation principles that occur in the study of Markov processes and come in a distinctly `probabilistic' flavor, see for example~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the `mathematics'\footnote{The `mathematics' vs `physics' distinction is just a caricature and is not intended to be read into or thought about too deeply :)} side by working with branching processes/SPDEs and using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators (semigroup theory), weak/mild solutions, and the martingale problem (See for example~\cite{etheridge_mathematical_2011} or~\cite{week_white_2021}). Rigorously establishing a relation between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and SPDEs of the form \eqref{functional_langevin} (or those studied in~\cite{week_white_2021}) for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object.

To the best of my knowledge, a general formulation of stochastic field equations for the population dynamics of quantitative traits from the functional Fokker-Planck side in the manner I have carried out here has not been done before. Similar equations have been formulated for some specific stochastic models of quantitative trait evolution~\citep{rogers_demographic_2012,rogers_modes_2015} and population ecology of size-structured communities~\citep{odwyer_integrative_2009}. Stochastic field equations are also known in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014}, and have recently been proposed in a model of collective motion~\citep{o_laighleis_minimal_2018}. Broadly similar deterministic field theoretic approaches have also been proposed to study bio-geography and spatial biodiversity patterns such as the species-area relationship~(\cite{odwyer_field_2010}; But see~\cite{grilli_absence_2012} for an explanation of why the calculations presented in~\cite{odwyer_field_2010} are not exactly correct\footnote{The essential problem is that an assumption about time reversibility (`detailed balance' in more accurate physics lingo), usually made for equilibrium systems, is not satisfied. Nevertheless, the general \emph{approach} is still interesting and seems reasonable}). The ubiquity of such approaches in diverse areas of theoretical biology suggests that the formal systematic study and analysis of such equations could have wide-spread applications. Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory or quantum field theory, and are largely attacked using ingenious (and often somewhat system-specific) arguments and tools that may not necessarily be mathematically well understood~(\cite{carmona_stochastic_1999}; but see \cite{bogachev_fokker-planck-kolmogorov_2015} for a recent rigorous treatment of some infinite-dimensional Fokker-Planck equations). Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - For example, a careful reading of the literature provides many hints and recent attempts illustrating how the major ideas behind the Fock space representation~\citep{del_razo_probabilistic_2022}, perturbative `loop diagrams/expansions'~\citep{dodd_many-body_2009}, and the path integral formalism~\citep{doi_second_1976, peliti_path_1985, chow_path_2015, weber_master_2017} can all be co-opted to coax biological insights from the sort of stochastic processes I formulate as a model for evolutionary dynamics via equation \eqref{disc_functional_field_eqns}.

Equation \eqref{disc_functional_field_eqns} thus provides a (relatively) more accessible alternative formulation that does not (explicitly) rely on tools from relatively `advanced' mathematical fields such as martingale theory and stochastic analysis that are typically used to study the evolution of quantitative traits in finite, fluctuating populations~\citep{dawson_stochastic_1975,fleming_measure-valued_1979,ethier_markov_1986,champagnat_unifying_2006,etheridge_mathematical_2011,week_white_2021}. This is intended to make the models and ideas more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis and encourage the use of techniques from physics and related domains, as outlined in the paragraphs above. Importantly, the formalism I develop here in terms of functional Fokker-Planck equations likely does \emph{not} carry over to the study of higher dimensional quantitative traits (or populations which vary in two or more one-dimensional quantitative traits) because these processes are routinely badly behaved in higher dimensions: In particular, a probability density $P(\phi,t)$ with respect to the reals frequently does not even exist in higher dimensions if one has any biologically non-trivial features such as interactions between types (\cite{fleming_measure-valued_1979, walsh_introduction_1986}; Also see for example~\cite{evans_measure-valued_1994}), rendering equation \eqref{disc_functional_field_eqns} entirely meaningless. My (admittedly limited) understanding is that such processes are also rather difficult to study in two or more dimensions from the SPDE side for similar reasons since one is forced to work with distribution valued (rather than function valued) solutions and thus needs to use a considerable amount of advanced functional analysis to establish even basic existence-uniqueness results~\citep{walsh_introduction_1986,carmona_stochastic_1999,prevot_concise_2007,liu_stochastic_2015,balan_gentle_2018}. It may well be the case that concrete biologically useful progress in this direction requires new mathematics altogether, a situation increasingly also encountered in other areas of mathematical biology~\citep{vittadello_open_2022}. Just as theoretical physics has done in the past, the pursuit of general mathematical descriptions of biological theory may therefore also inspire new ideas that are of independent interest to `pure' mathematicians~\citep{cohen_mathematics_2004}.

### 2023-03-12 ind_files/99_Appendix_A.tex
Minor phrasing changes, add some references

Since this is true for an arbitrary choice of $f(x)$ (as long as $f$ is $C^2$), we are thus led to conclude that the density function $P(x,t)$ must satisfy:

@@

Equation \eqref{FPE} is the Fokker-Planck equation in one dimension. Using the exact same strategy, the multidimensional Fokker-Planck equation for the $m$-dimensional It\^{o} Process $d\mathbf{X}_t = \boldsymbol{\mu}(\mathbf{X}_t,t)dt + \boldsymbol{\sigma}(\mathbf{X}_t,t)dB_t$ can be easily found to be:

@@

where $\mathbf{D} = \mathbf{\sigma}\mathbf{\sigma}^T$. As an aside, I briefly note that some authors like to define the `probability current' $\mathbf{J}(\mathbf{x},t)$, an $m$-dimensional function with $i\textsuperscript{th}$ element

>>>>

Since this is true for an arbitrary choice of $f(x)$, we are thus led to conclude that the density function $P(x,t)$ must satisfy:

@@

Equation \eqref{FPE} is the Fokker-Planck equation in one dimension. Using the exact same strategy, the multidimensional Fokker-Planck equation for the $m$-dimensional It\^{o} Process $d\mathbf{X}_t = \boldsymbol{\mu}(\mathbf{X}_t,t)dt + \boldsymbol{\sigma}(\mathbf{X}_t,t)d\mathbf{B}_t$ can be easily found to be:

@@

where $\mathbf{D} = \mathbf{\sigma}\mathbf{\sigma}^T$. Some authors like to define the `probability current' $\mathbf{J}(\mathbf{x},t)$, an $m$-dimensional function with $i\textsuperscript{th}$ element
